\pkg{cubble}: An \proglang{R} Package for Organizing and Wrangling Multivariate Spatio-temporal Data

### 2023-06-18 2-cubble.Rmd
section 2 and 3 done

<!-- Spatio-temporal data -->

<!-- remote sensing, spatio-temporal cube: longitude, latitude, and time -->

<!-- trajectory: each observation has unique coordinate (spatial) and temporal information -->

<!-- the spacetime paper classifies spatio-temporal data into four different categories based on different spatial and temporal characteristics.  -->

<!-- Analysts to be clear about the spatio-temporal layouts and choose the corresponding data structure for analysis. -->

<!-- Spatio-temporal data can be conceptualized using a cubical data model with three axes which typically are, time, latitude and longitude. This abstraction can be useful for generalizing operations and visualization purposes: @lu_multidimensional_2018 shows how array operations (select, scale, reduce, rearrange, and compute) can be mapped onto the cube; @bach_review_2014 reviews the temporal data visualization based on space-time cube operations. Notice that the term space-time cube in their article "does not need to involve spatial data", but refers to "an abstract 2D substrate that is used to visualize data at a specific time". Despite its main focus being on temporal data, the mindset of abstracting out data representation to construct visualizations, still applies to our spatio-temporal data manipulation and visualization approaches. -->

<!-- The most common space-time cube uses the three axes, time, latitude, longitude, and can be considered stacking space across time. Ours is a multivariate spatio-temporal cube with the three axes defined to be time, site and variables, as illustrated in the leftmost column of Figure \ref{fig:cube-def}. The time axis is the same in both versions, while the site axis now captures both latitude and longitude. Finally, variables are stacked on this space-time canvas, with one observation per site and time point. This notion is adopted to avoid using hyper-cubes when describing multivariate spatio-temporal data and is the conceptual framework behind the \code{cubble} class. With this conceptual model, operations on spatio-temporal data can be mapped to operations on the cube and the rest of Figure \ref{fig:cube-def} show examples of slicing on site, time, and variable.  -->

[conceptual framework]

Spatio-temporal data can encompass data with various spatial and temporal characteristics and different data require different structures for wrangling and analysis: climate weather stations record data at fixed point location but may suffer from potential temporal data quality issue such as missing data for certain days. GPS data tracks unique point locations at different timestamps and is represented as trajectories. Satellite imageries capture snapshots of landscapes at selected time and is commonly structured as raster data. The spatio-temporal data cubble address are those collected at unique fixed locations while allowing for irregularity in the temporal dimension, such as the weather station data. This corresponds to the full space-time and sparse space-time layouts in the spacetime paper [@spacetime].

@@

-   a nested/ spatial cubble is represented by the class `c("spatial_cubble_df", "cubble_df")`
-   a long/ temporal cubble is represented by the class `c("temporal_cubble_df", "cubble_df")`

@@

A cubble object inherits the attributes from tibble (and its subclasses): `class`, `row.names`, and `names`. Additionally, it has three specialised attributes:

-   `key`: the spatial identifier
-   `index`: the temporal identifier
-   `coords`: a pair of ordered coordinates associated with the location

Readers who are familiar with the `key` and `index` attributes from the `tsibble` package would already understand the two arguments. In cubble, the `key` attribute identifies the row in the nested cubble, and when combined with the `index` argument, it identifies the row in the long cubble. Currently, cubble only supports one variable as the key, and the accepted temporal classes for index includes the base R classes `Date`, `POSIXlt`, `POSIXct`, as well as tsibble's `yearmonth`, `yearweek`, and `yearquarter` classes.

The `coords` attribute represents an ordered pair of coordinates. It can be either an unprojected pair of longitude and latitude, or a projected easting and northing value. The `sf` package is used under teh hood to calculate the bounding box, displayed in the header of a nested cubble.

>>>>

Spatio-temporal data can encompass data with various spatial and temporal characteristics and different data require different structures for wrangling and analysis: climate weather stations record data at fixed point location but may suffer from potential temporal data quality issue such as missing data for certain days. GPS data tracks unique point locations at different timestamps and is represented as trajectories. Satellite imageries capture snapshots of landscapes at selected time and is commonly structured as raster data. The spatio-temporal data cubble address are those collected at unique fixed locations while allowing for irregularity in the temporal dimension, such as the weather station data. This corresponds to the full space-time and sparse space-time layouts described in the spacetime paper [@spacetime] and reproduced in Figure  \ref{fig:layout}.

```{r layout, echo = FALSE, fig.cap="The full space-time and sparse space-time layouts handled by the cubble package. Reproduced from Figure 1 in the spacetime paper \\citep{spacetime}.", fig.width = 8, fig.height=4, fig.align='center'}
library(tidyverse)
library(patchwork)
p1 <- tibble(
  id = 1:12,
  x = rep(1:4, each = 3),
  y = rep(1:3, 4)
) %>% 
  ggplot(aes(x = as.factor(x) ,y = as.factor(y))) + 
  geom_point() + 
  geom_text(aes(y = y - 0.2, label = id)) + 
  theme_bw() + 
  theme(aspect.ratio = 1) + 
  scale_x_discrete(labels = c("1st", "2nd", "3rd", "4th")) + 
  scale_y_discrete(labels = c("1st", "2nd", "3rd", "4th")) + 
  labs(x = "Time points", y = "Spatial features", title = "STF: full grid layout")

p2 <- tibble(
  id = 1:7,
  x = c(1, 1, 1, 2, 2, 3, 4),
  y = c(1, 2, 3, 2, 3, 1, 2),
  label = c("1[1,1]", "2[1,2]", "3[1,3]", "4[2,2]", "5[3,2]", "6[1,3]", "7[2,4]")
) %>% 
  ggplot(aes(x = as.factor(x) ,y = as.factor(y))) + 
  geom_point() + 
  geom_text(aes(y = y - 0.2, label = label)) + 
  theme_bw() + 
  theme(aspect.ratio = 1) + 
  scale_x_discrete(labels = c("1st", "2nd", "3rd", "4th")) + 
  scale_y_discrete(labels = c("1st", "2nd", "3rd", "4th")) + 
  labs(x = "Time points", y = "Spatial features", title = "STS: sparse grid layout")

p1 | p2
```

@@

-   a nested/ spatial cubble with class `c("spatial_cubble_df", "cubble_df")`
-   a long/ temporal cubble with class `c("temporal_cubble_df", "cubble_df")`

@@

A cubble object inherits the attributes from tibble (and its subclasses): `class`, `row.names`, and `names`. Additionally, it has three specialised attributes: `key`, `index`, and `coords`. Readers who are familiar with the `key` and `index` attributes from the `tsibble` package would already understand the two arguments. In cubble, the `key` attribute identifies the row in the nested cubble (given the use of `tidyr::nest()` for nesting), and when combined with the `index` argument, it identifies the row in the long cubble. Currently, cubble only supports one variable as the key, and the accepted temporal classes for index includes the base R classes `Date`, `POSIXlt`, `POSIXct`, as well as tsibble's `yearmonth`, `yearweek`, and `yearquarter` classes.

The `coords` attribute represents an ordered pair of coordinates. It can be either an unprojected pair of longitude and latitude, or a projected easting and northing value. The `sf` package is used under the hood to calculate the bounding box, displayed in the header of a nested cubble.

### 2023-06-18 3-advanced.Rmd
section 2 and 3 done

The workflow with the \code{cubble} class works well with an interactive graphics pipeline (e.g., @buja1988elements; @buja1996interactive;  @sutherland2000orca; @xie2014reactive; @cheng2016enabling) that is available in \proglang{R} with the package \pkg{crosstalk} [@crosstalk]. Figure \ref{fig:illu-interactive} illustrates how linking can be achieved between a map and multiple time series in a \code{cubble} object. The map (produced from the nested form) and time series (produced from the long form) are both shared \code{crosstalk} objects. When a user makes a selection on the map, the site is highlighted (a). This activates a row in the nested \code{cubble}, which is then communicated to the long \code{cubble} -- all the observations with the same id (b) will be selected. The long \code{cubble} will then highlight the corresponding series in the time series plot (c).

Linking is also available starting from the time series plot, by selecting points. This will activate rows having the same id in the long \code{cubble}. The corresponding rows in the nested \code{cubble} are activated, and highlighted on the map. (An illustration can be found in the appendix.) Note that this type of linking, both from the map or the time series, is what @CS07 would call categorical variable linking, where station id is the categorical variable.

>>>>

The workflow with the \code{cubble} class works well with an interactive graphics pipeline (e.g., @buja1988elements; @buja1996interactive;  @sutherland2000orca; @xie2014reactive; @cheng2016enabling) that is available in \proglang{R} with the package \pkg{crosstalk} [@crosstalk]. Figure \ref{fig:illu-interactive} illustrates the linking between a map and multiple time series in a \code{cubble} object. Both the map, generated from the nested cubble, and the time series plot, generated from the long cubble, are shared crosstalk objects. When a user selects a location on the map (a), the corresponding site is highlighted. This selection activates a row in the nested \code{cubble}, which is then communicated to the long \code{cubble}, resulting in teh selection of all observations with the same ID in (b). Consequently the long \code{cubble} highlights the corresponding series in the time series plot (c). Linking can also be initiated from the time series plot, by selecting points on the time series. This action activates rows with the same ID in the long \code{cubble} and the row in the nested \code{cubble}, which will be highlighted on the map. Please refer to the appendix for the corresponding illustration.

### 2023-06-18 jss/main.tex
section 2 and 3 done

{[}conceptual framework{]}

Spatio-temporal data can encompass data with various spatial and temporal characteristics and different data require different structures for wrangling and analysis: climate weather stations record data at fixed point location but may suffer from potential temporal data quality issue such as missing data for certain days. GPS data tracks unique point locations at different timestamps and is represented as trajectories. Satellite imageries capture snapshots of landscapes at selected time and is commonly structured as raster data. The spatio-temporal data cubble address are those collected at unique fixed locations while allowing for irregularity in the temporal dimension, such as the weather station data. This corresponds to the full space-time and sparse space-time layouts in the spacetime paper \citep{spacetime}.

@@

\begin{itemize}
\tightlist
\item
  a nested/ spatial cubble is represented by the class \texttt{c("spatial\_cubble\_df",\ "cubble\_df")}
\item
  a long/ temporal cubble is represented by the class \texttt{c("temporal\_cubble\_df",\ "cubble\_df")}
\end{itemize}

@@

A cubble object inherits the attributes from tibble (and its subclasses): \texttt{class}, \texttt{row.names}, and \texttt{names}. Additionally, it has three specialised attributes:

\begin{itemize}
\tightlist
\item
  \texttt{key}: the spatial identifier
\item
  \texttt{index}: the temporal identifier
\item
  \texttt{coords}: a pair of ordered coordinates associated with the location
\end{itemize}

Readers who are familiar with the \texttt{key} and \texttt{index} attributes from the \texttt{tsibble} package would already understand the two arguments. In cubble, the \texttt{key} attribute identifies the row in the nested cubble, and when combined with the \texttt{index} argument, it identifies the row in the long cubble. Currently, cubble only supports one variable as the key, and the accepted temporal classes for index includes the base R classes \texttt{Date}, \texttt{POSIXlt}, \texttt{POSIXct}, as well as tsibble's \texttt{yearmonth}, \texttt{yearweek}, and \texttt{yearquarter} classes.

The \texttt{coords} attribute represents an ordered pair of coordinates. It can be either an unprojected pair of longitude and latitude, or a projected easting and northing value. The \texttt{sf} package is used under teh hood to calculate the bounding box, displayed in the header of a nested cubble.

@@

\hypertarget{using-a-tsibble-for-the-temporal-component}{%
\subsubsection{Using a tsibble for the temporal component}\label{using-a-tsibble-for-the-temporal-component}}

@@

\hypertarget{using-sf-for-the-spatial-component}{%
\subsubsection{Using sf for the spatial component}\label{using-sf-for-the-spatial-component}}

@@

One common task when working with spatio-temporal data is to match nearby sites. For example, we may want to verify the location of an old list of stations with current stations, or we may want to match the data from different data sources. In cubble, spatial and temporal matching are performed using the functions \texttt{match\_spatial()} and \texttt{match\_temporal()}. The \texttt{match\_spatial()} function calculates the spatial distance between observations in two cubble objects. Various distance measures are available (check \texttt{sf::st\_distance}). Analysts can specify the number of matched groups to output using the \texttt{spatial\_n\_group} argument (default to 4 groups) and the number of matches per group using the \texttt{spatial\_n\_group}argument (default to 1, one-to-one matching). The syntax to use \texttt{match\_spatial()} is:

@@

The workflow with the \code{cubble} class works well with an interactive graphics pipeline (e.g., \citet{buja1988elements}; \citet{buja1996interactive}; \citet{sutherland2000orca}; \citet{xie2014reactive}; \citet{cheng2016enabling}) that is available in \proglang{R} with the package \pkg{crosstalk} \citep{crosstalk}. Figure \ref{fig:illu-interactive} illustrates how linking can be achieved between a map and multiple time series in a \code{cubble} object. The map (produced from the nested form) and time series (produced from the long form) are both shared \code{crosstalk} objects. When a user makes a selection on the map, the site is highlighted (a). This activates a row in the nested \code{cubble}, which is then communicated to the long \code{cubble} -- all the observations with the same id (b) will be selected. The long \code{cubble} will then highlight the corresponding series in the time series plot (c).

Linking is also available starting from the time series plot, by selecting points. This will activate rows having the same id in the long \code{cubble}. The corresponding rows in the nested \code{cubble} are activated, and highlighted on the map. (An illustration can be found in the appendix.) Note that this type of linking, both from the map or the time series, is what \citet{CS07} would call categorical variable linking, where station id is the categorical variable.

>>>>

Spatio-temporal data can encompass data with various spatial and temporal characteristics and different data require different structures for wrangling and analysis: climate weather stations record data at fixed point location but may suffer from potential temporal data quality issue such as missing data for certain days. GPS data tracks unique point locations at different timestamps and is represented as trajectories. Satellite imageries capture snapshots of landscapes at selected time and is commonly structured as raster data. The spatio-temporal data cubble address are those collected at unique fixed locations while allowing for irregularity in the temporal dimension, such as the weather station data. This corresponds to the full space-time and sparse space-time layouts described in the spacetime paper \citep{spacetime} and reproduced in Figure \ref{fig:layout}.

{\centering \includegraphics{/Users/hzha400/Documents/research/paper-cubble/figures/layout-1}

\caption[The full space-time and sparse space-time layouts handled by the cubble package]{The full space-time and sparse space-time layouts handled by the cubble package. Reproduced from Figure 1 in the spacetime paper \citep{spacetime}.}\label{fig:layout}
\end{figure}
\end{CodeChunk}

@@

\begin{itemize}
\tightlist
\item
  a nested/ spatial cubble with class \texttt{c("spatial\_cubble\_df",\ "cubble\_df")}
\item
  a long/ temporal cubble with class \texttt{c("temporal\_cubble\_df",\ "cubble\_df")}
\end{itemize}

@@

A cubble object inherits the attributes from tibble (and its subclasses): \texttt{class}, \texttt{row.names}, and \texttt{names}. Additionally, it has three specialised attributes: \texttt{key}, \texttt{index}, and \texttt{coords}. Readers who are familiar with the \texttt{key} and \texttt{index} attributes from the \texttt{tsibble} package would already understand the two arguments. In cubble, the \texttt{key} attribute identifies the row in the nested cubble (given the use of \texttt{tidyr::nest()} for nesting), and when combined with the \texttt{index} argument, it identifies the row in the long cubble. Currently, cubble only supports one variable as the key, and the accepted temporal classes for index includes the base R classes \texttt{Date}, \texttt{POSIXlt}, \texttt{POSIXct}, as well as tsibble's \texttt{yearmonth}, \texttt{yearweek}, and \texttt{yearquarter} classes.

The \texttt{coords} attribute represents an ordered pair of coordinates. It can be either an unprojected pair of longitude and latitude, or a projected easting and northing value. The \texttt{sf} package is used under the hood to calculate the bounding box, displayed in the header of a nested cubble.

@@

\hypertarget{using-a-tsibble-object-as-the-temporal-component}{%
\subsubsection{Using a tsibble object as the temporal component}\label{using-a-tsibble-object-as-the-temporal-component}}

@@

\hypertarget{using-an-sf-object-as-the-spatial-component}{%
\subsubsection{Using an sf object as the spatial component}\label{using-an-sf-object-as-the-spatial-component}}

@@

One common task when working with spatio-temporal data is to match nearby sites. For example, we may want to verify the location of an old list of stations with current stations, or we may want to match the data from different data sources. In cubble, spatial and temporal matching are performed using the functions \texttt{match\_spatial()} and \texttt{match\_temporal()}. The \texttt{match\_spatial()} function calculates the spatial distance between observations in two cubble objects. Various distance measures are available (check \texttt{sf::st\_distance}). Analysts can specify the number of matched groups to output using the \texttt{spatial\_n\_group} argument (default to 4 groups) and the number of matches per group using the \texttt{spatial\_n\_group} argument (default to 1, one-to-one matching). The syntax to use \texttt{match\_spatial()} is:

@@

The workflow with the \code{cubble} class works well with an interactive graphics pipeline (e.g., \citet{buja1988elements}; \citet{buja1996interactive}; \citet{sutherland2000orca}; \citet{xie2014reactive}; \citet{cheng2016enabling}) that is available in \proglang{R} with the package \pkg{crosstalk} \citep{crosstalk}. Figure \ref{fig:illu-interactive} illustrates the linking between a map and multiple time series in a \code{cubble} object. Both the map, generated from the nested cubble, and the time series plot, generated from the long cubble, are shared crosstalk objects. When a user selects a location on the map (a), the corresponding site is highlighted. This selection activates a row in the nested \code{cubble}, which is then communicated to the long \code{cubble}, resulting in teh selection of all observations with the same ID in (b). Consequently the long \code{cubble} highlights the corresponding series in the time series plot (c). Linking can also be initiated from the time series plot, by selecting points on the time series. This action activates rows with the same ID in the long \code{cubble} and the row in the nested \code{cubble}, which will be highlighted on the map. Please refer to the appendix for the corresponding illustration.
